Cheminformatics, the Computer Science of Chemical Discovery, Turning Open-Source

### 2012-02-06 resubmit/cheminformatics_CACM_revision.tex
Finished off second round

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (top left), a labeled graph (top right), a 3D
  conformation (bottom left) and a surface representation (bottom right). All show differing
  (and valid) aspects of the same molecule and are suited for
  different purposes, with the choice of representation guided by its
  intended use.}
\label{figure:chemical-structures}
\end{figure}

Although graphs implicitly encode the 3D structure of a molecule (when
combined with knowledge about favored bond angles and distances), many
different low energy 3D structures, or conformers, may be consistent
with the same graph structure. In addition, the 3D structure may have
a ``secondary'' geometrical arrangement of features (such as the
presence of a right-handed helix) which cannot be encoded in the
graph. Thus we have 3D representations
(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
goes further and represents a molecular surface, encoding some
property (such as lipophilicity, the ability to dissolve in non-polar
solvents) that varies over the surface.

Even though some structure representations contain more explicit
information than others, they are all equally valid and are suited for
different problems. Fig.~\ref{figure:chemical-structures} shows
a hierarchy of representations; when searching for
substructures, a 2D representation suffices, but when exploring
protein binding, a 3D structure is vital.

Given that a fundamental principle of the field is that \emph{similar
molecules exhibit similar properties} \cite{Johnson:1990qf}, the
choice of representation is key in determining how such similarities
are evaluated, and thus the effectiveness of subsequent analyses. But
even after one has selected a representation one faces the challenge
of balancing computational costs with the utility of the
representation. For example, a full 3D description of a molecule,
taking into account all possible conformers would let us accurately
predict many properties. But the size of this representation and the
time required to evaluate it would be prohibitive. Instead, can we
obtain accurate predictions from a subset of conformers? Or, can we
obtain comparable accuracy by using a 2D representation? And if so,
what type of labels are required? Currently, many of these questions
are answered using trial and error.

@@

@@

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. Here
multiple approaches exist. We can transform molecular graphs into a
numerical vector representation of chemical structures
\cite{todeschini2000}, ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measured similarities between the
graph's nodes. In these cases, rather than compute vector
representations, we directly operate on the graph representations.

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we
looking for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
also some explanation as to why a molecule is predicted as toxic or
non-toxic (for example)? The former scenario is common in virtual
screening settings, where we require a high degree of accuracy and fast
predictions, but do not really care why one molecule is active
and another inactive. These types of models can be black boxes (e.g., neural
networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we need to understand the effect of a
certain structural feature on the observed potency and expect that the
model will provide insight. These types of models are generally
distributional in nature (such as linear regression and Partial Least Squares) though
some algorithmic approaches such as random forests can provide insight.

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation.
Traditionally, predictive models have focused on a single
structure for a small molecule and ignored the receptor. Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that the 3D geometries of that molecule be taken
into account and, as far as possible, that the receptor also be considered.
Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger \textit{et al.}
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

@@

\begin{figure}
\centering
\includegraphics[height=1.3in]{platins.png}
\caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
\label{figure:cistrans}
\end{figure}
%
\begin{figure}
\centering
\includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
\caption{Illustration of the ``Applicability Domain'' concept. In order to estimate whether model predictions are reliable, it is crucial to define areas of ``chemical space'' where the model is applicable, and where it is not. In this case, the ``model domain'' includes the molecule at the top. The (relatively similar) molecules to the left are likely to be included in the Applicability Domain of the model, while the (more dissimilar) molecules to the right are likely located outside this domain, hence the generated model is probably not applicable.}
\label{figure:applicability-domain}
\end{figure}
%
\end{document}

>>>>

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (top left), a labeled graph (top
  right; 'A' corresponds to aromatic atoms, 'O' to oxygens and 'On' to
  oxygens or nitrogens. Unlabeled nodes have an implicit label of 'C'), a 3D
  conformation (bottom left) and a surface representation (bottom right). All show differing
  (and valid) aspects of the same molecule and are suited for
  different purposes, with the choice of representation guided by its
  intended use.}
\label{figure:chemical-structures}
\end{figure}

Although graphs implicitly encode the 3D structure of a molecule (when
combined with knowledge about favored bond angles and distances), many
different low energy 3D structures, or conformers, may be consistent
with the same graph structure.

In addition, the 3D structure may have
a ``secondary'' geometrical arrangement of features (such as the
presence of a right-handed helix) which cannot be encoded in the
graph. Thus we have 3D representations
(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
goes further and represents a molecular surface, encoding some
property (such as lipophilicity, the ability to dissolve in non-polar
solvents) that varies over the surface.

Even though some structure representations contain more explicit
information than others, they are all equally valid and are suited for
different problems. Fig.~\ref{figure:chemical-structures} shows a
hierarchy of representations; when searching for substructures, a 2D
representation suffices, but when exploring protein binding, a 3D
structure is vital. Figure \ref{fig:structure-classes} in the Appendix
highlights yet another subtlety where the number and types of atoms
and their connectivity may not suffice to uniquely define the
structure.

Given that a fundamental principle of the field is that \emph{similar
molecules exhibit similar properties} \cite{Johnson:1990qf}, the
choice of representation is key in determining how such similarities
are evaluated, and thus the effectiveness of subsequent analyses. But
even after one has selected a representation one faces the challenge
of balancing computational costs with the utility of the
representation. For example, a full 3D description of a molecule,
taking into account all possible conformers would let us accurately
predict many properties. But the size of this representation and the
time required to evaluate it would be prohibitive. Instead, can we
obtain accurate predictions from a subset of conformers? Or, can we
obtain comparable accuracy by using a 2D representation? And if so,
what type of labels are required? Currently, many of these questions
are answered using trial and error - in that an objective function
(usually root mean square error or percentage correct) is defined and
descriptors and modeling approaches are iteratively adapted to
optimize the objective function . In other words, there is no
underlying theory to explain or suggest optimal approaches.

@@

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. For
example, labeled graphs (Figure 1B), and their associated
characterizations are easily accessible to computer scientists, yet
such features miss significant physicochemical features (surface
properties such as charge, flexibility and so on). At the same time,
it can be difficult to objectively quantify many chemical aspects of a
molecule, such that the resultant descriptors are suitable for
predictive modeling.  Hence, the choice of a chemical descriptor
should by no means be treated as a ‘solved’ problem. For a more
detailed discussion the reader is referred to more comprehensive
textbooks \cite{todeschini2000,faulon2010}.

As noted above, molecular graphs can be transformed to numerical
vector representations ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measured similarities between the
graph's nodes. In these cases, rather than compute vector
representations, we directly operate on the graph representations.

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we looking
for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
also some explanation as to why a molecule is predicted as toxic or
non-toxic (for example)? The former scenario is common in virtual
screening settings, where we require a high degree of accuracy and
fast predictions, but do not really care why one molecule is active
and another inactive. These types of models can be black boxes (e.g.,
neural networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we need to understand the effect of a certain
structural feature on the observed potency and expect that the model
will provide insight. These types of models are generally
distributional in nature (such as linear regression, Partial Least
Squares and Na\"{i'}ve Bayes) though some algorithmic approaches such
as random forests can provide insight.

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation.
Traditionally, predictive models have focused on a single structure
for a small molecule and ignored the receptor. Yet, small molecules
can exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that the 3D geometries of that molecule be taken
into account and, as far as possible, that the receptor also be
considered. While some relevant low-energy conformers of small
molecules may be accessible from crystallographic databases, this is
not always the case. Though it is now possible to generate reasonable
low energy conformations \emph{ab initio} it must be noted that the
``biologically relevant'' conformation might differ from the lowest
energy conformation of the molecule considered in isolation,
necessitating the need for conformational search.  Multi-conformer
modeling has been addressed in the 4D-QSAR methodology described by
Hopfinger \textit{et al.}  \cite{Albuquerque:1998ys}.  Recent
techniques such as \emph{multiple-instance learning} could also be
applied to the multi-conformer problem.

@@

@@

\begin{figure}
\centering
\includegraphics[height=1.3in]{platins.png}
\caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
\label{figure:cistrans}
\end{figure}
%
\begin{figure}
\centering
\includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
\caption{Illustration of the ``Applicability Domain'' concept. In order to estimate whether model predictions are reliable, it is crucial to define areas of ``chemical space'' where the model is applicable, and where it is not. In this case, the ``model domain'' includes the molecule at the top. The (relatively similar) molecules to the left are likely to be included in the Applicability Domain of the model, while the (more dissimilar) molecules to the right are likely located outside this domain, hence the generated model is probably not applicable.}
\label{figure:applicability-domain}
\end{figure}
%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{cacm_tautomers.png}
\caption{An depiction of the various classes of chemical structures
  (shown using a 2D representation). This figure serves to highlight
  the danger of assuming that a molecule is \emph{uniquely} defined by
the number and types of atoms and even simple connectivity between them.}
\label{fig:structure-classes}
\end{figure}
%

\end{document}
